---
title: "Water Quality Predictions"
author: "Bridget Gervais"
date: "2026-06-25"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Welcome to this page on Machine learning predictions of water quality. Nitrate is an incredibly important issue, especially in agriculturally driven states. Excess fertilizer from the soil runs off into waterways and causes health issues, thus predicting high nitrates can help offset treatment costs and reduce future runoff. 

The below map allows you to click on a site and on the right, a plot of the 4-year monthly nitrate trend will appear. Each color denotes a region that was used to subset the data into hydrologically similar sites. The regions are: 

- Mississippi Plain (MSP)
- Iowa Surface (IS)
- Des Moines Lobe (DML)
- Northwest Iowa Plain (NWIP)
- Missouri Plain (MOP)
- Southern Iowa Drift Plain (SIDP)

Feel free to explore!! 

```{r data preparation, echo=FALSE, warning = FALSE, message=FALSE}
library(shiny)
library(bslib)
library(ggmap)
library(ggplot2)
library(plotly)
library(dplyr)
library(tidyr)
library(DT)
library(leaflet)
library(MLmetrics)
library(e1071)

setwd("C:/Users/Bridget/Documents/Indiana_University/Spring _26/Senior Thesis/Journal Article Submission and Code")
nitrate <- read.csv("data.csv")

#assign region order
nitrate$Region <- factor(nitrate$Region, 
                         levels=c("NWIP","DML","MSP","MOP", "SIDP","IS"))

regions <- levels(nitrate$Region)

nitrate <- drop_na(nitrate)

nitrate$Warning_level <- factor(nitrate$Warning_level, 
                                levels = c("Safe","Warning","Alert"))

nitrate$Future_Warning <- factor(nitrate$Future_Warning, 
                                 levels = c("Safe","Warning","Alert"))

nitrate$MonthLabel <- factor(
  nitrate$Month,
  levels = 1:12,
  labels = month.abb
)
```

```{r, echo=FALSE}
# Color palette
pal <- colorFactor(
  palette = c("red","blue","green","gold","purple","magenta"),
  domain = levels(nitrate$Region)
)

#create page dimensions and items
ui <- fluidPage(
  titlePanel("Nitrate Warning - December 2025"),
  fluidRow(
    column(6,
        leafletOutput("map", height = "300px")),

    column(6,
        h3(textOutput("siteName")),
        plotlyOutput("trend", height = "300px"))))
```

```{r, echo=FALSE}
server <- function(input, output, session){
  #output map
output$map <- renderLeaflet({
  leaflet(nitrate) %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    setView(lng = -93.5,
            lat = 42.1,
            zoom = 6) %>%
    addCircleMarkers(lng = ~longitude.x,
                     lat = ~latitude.x,
                     layerId = ~Water_site,
                     color = ~pal(Region),
                     radius = 5,
                     stroke = TRUE,
                     fillOpacity = .85,
                     popup = ~paste("<b>", Water_site, "</b><br>",
                   "Region: ", Region))

})

#define selected site by click linking to layerID data
selectedSite <- reactive({
  req(input$map_marker_click)
  input$map_marker_click$id
})

#output text for selected site
output$site <- renderText({
  selectedSite()
})

#create trend object
output$trend <- renderPlotly({
  site_data <- nitrate %>%
    filter(Water_site == selectedSite())
  
  #x is month, y is nitrate, written in html
  plot_ly(site_data,
          x = ~MonthLabel,
          y = ~Avg_Nitrate,
          color = ~factor(Year),
          colors = "Dark2",
          type = "scatter",
          mode = "lines+markers",
          hovertemplate = paste("<b>Year %{fullData.name}</b><br>",
                                "%{x}<br>",
                                "Nitrate: %{y:.2f}<extra></extra>")) %>%
    
    layout(title = "Monthly Nitrate",
           xaxis = list(title = "Month"),
           yaxis = list(title = "Average Nitrate"))
})
}

shinyApp(ui, server)

```

## ML Model Overview

Models used include polynomial regression and support vector regression, using temperature and monthly rainfall as predictors, and nitrate change as the response. 

Check the box next to the region you want to view

```{r train-test, echo=FALSE}
#create train/test splitting function
define_train_test <- function(){
  train <- list()
  test <- list()
  
  #assign train and test set based on year
  for (i in 1:6){
    train[[i]] <- nitrate[nitrate$Year != 2025 & nitrate$Region == regions[i],]
    
    test[[i]] <- nitrate[nitrate$Year == 2025 & nitrate$Region == regions[i],]
  }
  return (list(train,test))
}

#create train and test variables
train_test <- define_train_test()
train <- train_test[[1]]
test <- train_test[[2]]

svm_params <- data.frame(sigma = c(.1, .1, .01, .1, .1, 1),
                         C = c(10, 10, 100, 10, 10, 1))

lm_models <- list()
svm_models <- list()
lm_predicted_change <- list()
svm_predicted_change <- list()

#train polynomial model via bootstrap
for (i in 1:6){
  set.seed(35791)
  lm_models[[i]] <- lm(Change~0+poly(Monthly_Rainfall,degree=2)+
                         poly(Avg_temp, degree=2), data = train[[i]])

  lm_predicted_change[[i]] <- predict(lm_models[[i]], test[[i]])
}

for (i in 1:6){
  set.seed(35791)
  svm_models[[i]] <- svm(Change ~ 0+Monthly_Rainfall+Avg_temp, 
                        data = train[[i]], 
                        kernel = "radial",
                        cost = svm_params$C[i],
                        gamma = svm_params$sigma[i], 
                        scale = TRUE,
                        type = 'eps-regression')

  svm_predicted_change[[i]] <- predict(svm_models[[i]], test[[i]])
}

#---Evaluate Model Performance
get_evaluators <- function(model, predicted_change){
  pred_warning <- list() #predicted warning level (Safe, Warning, Alert)
  predicted_outputs <- list() #expected nitrate concentration (mg/L) -- change + prev month
  resids <- list() #residuals of expected - actual change
  f1s <- list()
  recalls <- list()
  accuracy <- c()
  
  #iterate over each region
  for (i in 1:6){
    #Note: Remove nulls, which have been filtered to only be missing change variable, meaning
    #a prediction would not be able to be compared to the row, thus not used for testing. 
    #compute actual nitrate output
    predicted_outputs[[i]] <- test[[i]]$Avg_Nitrate + predicted_change[[i]] 
    
    #convert to warning class --> some low instances ended up less than zero, thus defaulted to safe class with min as the lower endpoint
    pred_warning[[i]] <- cut(predicted_outputs[[i]],breaks = c(-Inf,7.5,10,Inf), labels = c("Safe","Warning","Alert"),right = FALSE)
    
    #compute residuals
    resids[[i]] <- test[[i]]$Change - predicted_change[[i]]
    
    #f1s
    f1_low <- F1_Score(y_true = test[[i]]$Future_Warning, y_pred = pred_warning[[i]],
                       positive = "Safe")
    f1_med <- F1_Score(y_true = test[[i]]$Future_Warning, y_pred = pred_warning[[i]],
                       positive = "Warning")
    f1_high <- F1_Score(y_true = test[[i]]$Future_Warning, y_pred = pred_warning[[i]],
                        positive = "Alert")
    
    #recalls
    recall_low <- Recall(y_true = test[[i]]$Future_Warning, y_pred = pred_warning[[i]],
                         positive = "Safe")
    recall_med <- Recall(y_true = test[[i]]$Future_Warning, y_pred = pred_warning[[i]],
                         positive = "Warning")
    recall_high <- Recall(y_true = test[[i]]$Future_Warning, y_pred = pred_warning[[i]],
                          positive = "Alert")
    
    #accuracy
    accuracy[i] <- Accuracy(y_true = drop_na(test[[i]])$Future_Warning, y_pred = pred_warning[[i]])
    
    f1s[[i]] <- c(f1_low, f1_med, f1_high)
    recalls[[i]] <- c(recall_low, recall_med, recall_high)
    
  }
  return (list(f1s, recalls, accuracy, predicted_outputs, pred_warning, resids))
}

#initialize evaluators for linear model
linear_evals <- get_evaluators(lm_models, lm_predicted_change)
linear_f1s <- linear_evals[[1]]
linear_recalls <- linear_evals[[2]]
linear_accuracy <- linear_evals[[3]]
linear_predicted <- linear_evals[[4]]
linear_pred_warning <- linear_evals[[5]]
linear_resids <- linear_evals[[6]]

#initialize evaluators for SVM model
svm_mod <- get_evaluators(svm_models, svm_predicted_change)
svm_f1s <- svm_mod[[1]]
svm_recalls <- svm_mod[[2]]
svm_accuracy <- svm_mod[[3]]
svm_predicted <- svm_mod[[4]]
svm_pred_warning <- svm_mod[[5]]
svm_residuals <- svm_mod[[6]]

#fill in f1 nulls to 0 
for (i in 1:6){
  for (j in 1:3){
    if (is.na(linear_f1s[[i]][j])){
      linear_f1s[[i]][j] <- 0
    }
    if (is.na(svm_f1s[[i]][j])){
      svm_f1s[[i]][j] <- 0
    }
    
  }
}

```


```{r, echo=FALSE}
for (i in 1:6){
  test[[i]]$Linear_Predicted <- linear_predicted[[i]]
  test[[i]]$SVM_Predicted <- svm_predicted[[i]]
}

full_predictions <- rbind(test[[1]],test[[2]], test[[3]],test[[4]],test[[5]], test[[6]])

full_predictions$Linear_Warning <- cut(full_predictions$Linear_Predicted,
                                       breaks=c(-Inf,7.5,10,Inf),
                                       labels = c("Safe","Warning","Alert"))

full_predictions$SVM_Warning <- cut(full_predictions$SVM_Predicted,
                                       breaks=c(-Inf,7.5,10,Inf),
                                       labels = c("Safe","Warning","Alert"))

ui2 <- fluidPage(
  titlePanel("Regions"),
  sidebarLayout(
    sidebarPanel(
      checkboxGroupInput(
        inputId = "regions", 
        label = "Select Region(s):",
        choices = levels(full_predictions$Region), 
        selected = levels(full_predictions$Region))),
    mainPanel(
      fluidRow(
        column(6,plotlyOutput("modelPlotLM", height = "300px")),
        column(6,plotlyOutput("modelPlotSVM", height = "300px")))
      ), 
  ), DTOutput("metricsTable"))

server2 <- function(input, output, session){
  selectedRegions <- reactive({
    req(input$regions)
    
    full_predictions %>% filter(Region %in% input$regions)
  })
  
  output$modelPlotLM <- renderPlotly({
    region_data <- selectedRegions()
    
    plot_ly(region_data, 
            x= ~Future_Nitrate, 
            y= ~Linear_Predicted, 
            type = "scatter",
            mode = "markers", 
            color = ~factor(Year),
            colors = "Dark2",
            text = ~paste("<b>",Water_site,"</b><br>",
                          "Region: ", Region, "<br>",
                          "Month: ", MonthLabel, "<br>",
                          "Observed: ", Future_Nitrate, "mg/L <br>",
                          "Predicted: ", round(Linear_Predicted, 2), "mg/L"),
            hoverinfo = "text") %>%
      
      layout(title = "Model Predictions - Poly",
             xaxis = list(title = "Observed Nitrate"),
             yaxis = list(title = "Predicted Nitrate"), 
             color = "black",
             shapes = list(
               list(type = "line",
                    x0 = 0,
                    x1 = 25,
                    y0 = 0,
                    y1 = 25,
                    line = list(color = "blue", dash = "dash")),
               list(type = "line",
                    x0 = 0, 
                    x1 = 30, 
                    y0 = 10, 
                    y1 = 10, 
                    line = list(color = "red")),
               list(type = "line",
                    x0 = 0, 
                    x1 = 30, 
                    y0 = 7.5, 
                    y1 = 7.5, 
                    line = list(color = "gold"))), 
             annotations = list(
               list(x = 25, 
                    y = 8.3, 
                    text = "Warning", 
                    showarrow = FALSE,
                    font = list(size = 12)), 
               list(x = 23.5, 
                    y = 10.9, 
                    text = "Alert",
                    showarrow = FALSE, 
                    font = list(size=12))))
})
  
   output$modelPlotSVM <- renderPlotly({
    region_data <- selectedRegions()
    
    plot_ly(region_data, 
            x= ~Future_Nitrate, 
            y= ~SVM_Predicted, 
            type = "scatter",
            mode = "markers", 
            color = ~factor(Year),
            colors = "Dark2",
            text = ~paste("<b>",Water_site,"</b><br>",
                          "Region: ", Region, "<br>",
                          "Month: ", MonthLabel, "<br>",
                          "Observed: ", Future_Nitrate, "mg/L <br>",
                          "Predicted: ", round(SVM_Predicted, 2), "mg/L"),
            hoverinfo = "text") %>%
      
      layout(title = "Model Predictions - SVM",
             xaxis = list(title = "Observed Nitrate"),
             yaxis = list(title = "Predicted Nitrate"), 
             color = "black",
             shapes = list(
               list(type = "line",
                    x0 = 0,
                    x1 = 25,
                    y0 = 0,
                    y1 = 25,
                    line = list(color = "blue", dash = "dash")),
               list(type = "line",
                    x0 = 0, 
                    x1 = 30, 
                    y0 = 10, 
                    y1 = 10, 
                    line = list(color = "red")),
               list(type = "line",
                    x0 = 0, 
                    x1 = 30, 
                    y0 = 7.5, 
                    y1 = 7.5, 
                    line = list(color = "gold"))), 
             annotations = list(
               list(x = 25, 
                    y = 8.3, 
                    text = "Warning", 
                    showarrow = FALSE,
                    font = list(size = 12)), 
               list(x = 23.5, 
                    y = 10.9, 
                    text = "Alert",
                    showarrow = FALSE, 
                    font = list(size=12))))
})
  output$metricsTable <- renderDT({
    region_info <- selectedRegions()
    
    linear_rmse <- round(sqrt(mean((region_info$Linear_Predicted - region_info$Future_Nitrate)^2)),3)
    
    svm_rmse <- round(sqrt(mean((region_info$SVM_Predicted - region_info$Future_Nitrate)^2)),3)
    
    linear_accuracy <- paste0(round(mean(region_info$Linear_Warning == region_info$Future_Warning),3)*100,"%")
    
    svm_accuracy <- paste0(round(mean(region_info$SVM_Warning == region_info$Future_Warning),3)*100,"%")
    
    f1s_linear <- c()
    f1s_svm <- c()
    warnings <- c("Safe","Warning", "Alert")
    
    for (i in 1:3){
      f1s_linear[i] <- round(F1_Score(region_info$Future_Warning,
                           region_info$Linear_Warning, 
                           positive = as.character(
                             levels(region_info$Linear_Warning)[i])),3)
      
      f1s_svm[i] <- round(F1_Score(region_info$Future_Warning,
                          region_info$SVM_Warning,
                          positive = as.character(
                            levels(region_info$SVM_Warning)[i])),3)
    }

    
    datatable(data.frame(Model = c("Linear","SVM"),
                         RMSE = c(linear_rmse, svm_rmse),
                         Accuracy = c(linear_accuracy, svm_accuracy),
                         `F1 Safe` = c(f1s_linear[1], f1s_svm[1]),
                         `F1 Warning` = c(f1s_linear[2], f1s_svm[2]),
                         `F1 Alert` = c(f1s_linear[3], f1s_svm[3])), 
              rownames = FALSE,
              options = list(scrollY = "250px",
                             scrollCollapse = TRUE,
                             paging = FALSE))
  })
}


shinyApp(ui2, server2)
```

## Results - Discussion

The findings show how climate effects can be used to predict the expected change in nitrate concentration. The difficulty in predicting warnings is that the margin for the warning class is much smaller than the other classes. Any small deviation in the expected change, can put the new nitrate concentration into the wrong class. Likewise, the extremities between years added additional complexity to the model, making it unable to perform on the testing set. 

F1 scores are naturally large for classes where the majority of observations belong. While F1 scores work well for imbalanced datasets, it is important to note that the model didn't do better at predicting the safe class, but rather there are a lot more instances predicted correctly, making the precision and recall used for F1 scores higher. 








